Концентрация юрфирм и доступность юруслуг: как они связаны?

### 2024-05-18 legal-companies-concentration-and-availability/journal-article/Article.Rmd
Proofread of materials and methods, results and discussion, conclusion; add three references to introduction, #2

This paper empirically examines the relationship between the number of law firms in Russian cities and the provision of legal services. The existing literature on the availability of services typically focuses on medical services [@Joseph1982], with relatively little attention paid to legal services. Nevertheless, there is a growing recognition that spatial aspects of legal services, though understudied by geographers, are worthy of investigation [@Blacksell1988; @Patel2008]. This is because they are directly related to the rule of law, the protection of human rights, and the support of social order and stability. The analysis is based on the open data about small and medium-sized businesses (SMB) published by the Federal Tax Service (FTS) of Russia. The primary research question is how the specific provision of legal professionals is influenced by the number of lawyers. Additionally, the study aims to identify geographical patterns in the spatial distribution of legal services provision.

@@

**Data.** This research utilizes a geo-referenced dataset on small and medium-sized law firms incorporated in Russia. The dataset is created from FTS open data dumps (archives) of the state SMB Registry[^1] and FTS open data on the average number of employees of ogranizations[^2] using the Python command-line tool *ru-smb-companies*[^3] developed by the author of the paper. It contains information about the particular companies, including company's name, taxpayer ID, region and settlement of incorporation, main activity code. Law firms are selected by the main activity code equal to 69.10, and all the other companies were dropped from the source data. This technique of selection by the main activity code was derived from the analytical paper on the Russian legal services market prepared by the members of the Institute for the Rule of Law [@Moiseeva2016]. Additionally, the dataset was filtered by the following conditions: the year is 2021, the company is an organization rather than a sole entrepreneur, the company has declared non-zero revenue and expenditure for the target year (implying that the company is active), and the company's address of registration is in a city rather than in a rural area. In addition to the main dataset, a data on cities population freely distributed by the commercial company HFLabs is used [^4].

[^1]: https://www.nalog.gov.ru/opendata/7707329152-rsmp/
[^2]: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/
[^3]: https://github.com/PavelSyomin/russian-smb-companies
[^4]: https://github.com/hflabs/city

During the exploratory data analysis, six cities are identified as outliers subsequently excluded from the data. Two of these cities, Moscow and Saint Petersburg, are excluded due to the extremely high number of law firms employees: almost 18,866 and 5,550 respectively, while the next city has only less than 2,000 employees. The four other cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) are characterized by the very high proportion of law firms employees in city's population (from 38 per 10,000 in Krasnoarmeisk to 417 in Innopolis) compared to less than 20 per 10,000 in all the other cities.

**Methods.** A linear regression is the main method of data analysis used in this study. An independent variable for the regression is the number of employees of law firms in a city denoted as *n*. Theoretically, both number of law firms or their employees may be used, but law firms in different cities may have different size, thus the number of employees seems to be more stable and consistent metric. The value for each city is calculated as a sum of numbers of employees of each law firm incorporated in this city. If a law firm reported zero count of employees, it is assumed that the number of employees is equal to one. The dependent variable in the regression is specific provision of legal services which is denoted as *p* and calculated by dividing the number of employees of law firms in the city by the city's population and multiplying the result by 100,000. In other words, it's a number of law firms employees per 100,000 of population. This formula is a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016]. In addition, a grouping variable *g* categorizing the cities by population is used to check whether the overall relationship is dependent on the city size. The grouping by population is performed on the basis of the classification system of G. M. Lappo [-@Lappo1997]. Two linear models are fitted, one without grouping and one with grouping. The results are visialized on a plot. The analysis was performed using the R programming language and additional packages. The source code for this paper can be found in the public repository[^5].

[^5]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability

**Theoretical model.** Despite the perceived simplicity of the modelling described earlier, it has a theoretical mathematical background, and several possible types of outcome may be laid out right now before the analysis of the data in hand. It is known that a linear model is described by the two coefficients, intercept (a free term of the linear equation showing the relative shift of the regression line along the y axis) and slope (a coefficient before the independent variable showing the angle between the regression line and x axis). The slope can be either positive, negative, or zero. As a special edge case, it can be also infinite (i. e. regression line is vertical), even though, strictly speaking, such a dependency is not a function. Thus, the linear fit described earlier can yield a result of one of four types, each corresponding to one of the slope options and, subsequently, the pattern of relationship between the number of law firms and provision of legal services. These basic theoretical types are shown in Figure 1.

```{r figure-1-theoretical-models, fig.cap="**Figure 1** — Four plots showing four types of the theoretical relationship between the number of law firms employees and provision of lwgal services in cities", fig.height=3}
theoretical_models_plot
```

Type 1 corresponds to the zero slope. In this case, the regression line is horizontal. From the substantial point of view, it describes the absolute equality in the provision of legal services among cities, that is, the provision is independent of the number of law firms employees. Despite being quite unlikely, this type also may be described as optimal, because it guarantees equitable access to legal services (at least in spatial terms). Probably this type requires some sort of governmental intervention to form (e. g. encouraging the establishment of law firms in small cities).

Type 2 is characterized by the positive slope and, consequently, positive linear relationship between the number of law firms and the provision of legal services. It seems that at the free market this type is the most probable, even though the specific value of the slope may vary. In such case, the bigger the city, the easier it is to find a lawyer, and vise versa.

Type 3 is the reverse of Type 2: the slope of the regression line is negative, and so is the relationship between the number of law firms and the provision of legal services. It is also quite improbable. One possible situation in which this type may form is the unexpected and rapid migration of people from the smallest cities to the largest with almost uniform (Type 1) relationship as a prerequisite. Under such circumstances, the relative number of law firms in small cities increases fast, but the relative number of law firms in large cities can not grow immediately as fast as the population, and the largest cities start to suffer from the lack of legal services, while the smallest cities are oversaturated with law firms.

Type 4 is purely theoretical, but important in conjunction with the three other types. As Type 1, it also signals about the independence between the number of law firms and provision of legal services, but here the number of law firms employees is fixed rather than their proportion in the population. Again, it is very unlikely to happen in the real world, but it also suggests that some sort of governmental intervention (e. g. strictly controlling the number of law firms) is taking place.

@@

Figure 2 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in the Materials and methods section. The overall connection between the number of law firms and provision of legal services appears to be nonlinear and corresponds to a square root formula. However, the same relationship within each city size group is fitted well with a linear models. The coefficients of these models are different, and the slope decreases when the city population increases. While the regression lines for the small and medium cities are close to Type 4 (provision is independent of the number of lawyers because of the small variation in the number of lawers), the modelled relationship in millionaire cities resembles Type 1 (another option for independence, but due to relatively uniform and equal provision regardless the number of lawyers). The cities from big to extra-large are characterized by the positive relationship between the number of lawyers and the provision of legal services and can be classified as Type 2.

@@

```{r fig-2-empl-provision, fig.cap="**Figure 2** — The relationship between the number of lawyers and provision of legal services", fig.height=5}
ep_plot
```

In general, regional centers have relatively high number of lawyers and considerable level of legal services provision. This observation is probably explained by the centralization tendencies in the regions and the concentration of business and services in their administrative centers that situate the majority of government bodies and are also usually the most populous cities within a region.

Moscow and Saint Petersburg, though excluded from the data analysis and not shown on the main figure, are two remarkable outliers with extremely high concentration of legal services and a good level of legal services availability. Moscow alone accounts for almost one-third of all legal companies' employees, and Saint Petersburg accounts for about 10%. The position of the corresponding data points relative to the main regression lines suggests that there is an “availability saturation” level, where further concentration has little effect on availability.

The general tendencies observed in the Figure 2 should be supplemented by the analysis of the source data distribution shown in Figure 3 that extends the Figure 2 and displays the binned count of cities according to the number of lawyers and the provision of legal services. The overwhelming majority of cities is characterized by low level of both legal services provision and law firms development. Only a limited number (18) of cities is characterized by the high (more than 500) number of lawyers. It should be noted that high number of lawyers is good for legal services provision because all these cities are at least at the medium level of provision, and none of them has less than 5 lawyers per 10,000 people. However, cities with relatively low number of laywers can nevertheless reach top performance from the point of legal services provision that is on par with the cities with the highest number of lawyers. Moreover, it appears that such small yet well-served cities are even more typical than cities with high level of both lawyers and provision. The relative location of bin counts in the Figure 2 implies that there — besides the dominant bottom left corner — there are two main clusters of cities showing two different approaches of reaching high level of legal services provision. One approach is increase both in absolute number of law firms and relative number of lawyers per population probably charasteristic of the largest cities. Another is a relatively small yet sufficient number of lawyers typical probably for small or medium cities. Thus, apart from the general tendencies, the two-ways structure is found out.

```{r figure-3-empl-provision-binned, fig.cap="**Figure 3** — Count and proportion of cities with variuos combinations of number of lawyers and level of legal services provision"}
ep_binned_plot
```

Figure 4 is a map displaying the location of cities with relatively high (more than 5 lawyers per 10,000 people) provision of legal services. The most prominent cluster is formed by Moscow and its satellite cities. Other notable centers with high availability include Krasnodar and Stavropol in the South of Russia, Yekaterinburg and Chelyabinsk in the Urals, and Tuymen and Novosibirsk in Western Siberia. There are several regions with no cities of good legal services provision (especially on Far East, North, and Caucasus), and the majority of cities with relatively high level of legal services provision is located in the Central and Volga economic regions of the country.

@@

```{r fig-4-map, fig.cap="**Figure 4** — Legal services provision in Russian cities on the country's map. Only cities with more than 5 lawyers per 10,000 people are shown", fig.height=4}
map
```

@@

The provision of legal services in cities has a positive relationship with the number of lawyers. The overall trend can be fitted with a square root curve, while the relationship in distinct city size groups is close to linear with varying slope of the regression line. In small and medium cities (with population below 50,000), the increase of number of lawyers has little effect on the level of legal services provision, because the number of lawyers per 10,000 people varies greatly. In larger cities, the relationship is linear and positive. In the largest cities having the population above 1 million people, the impact of the growth of the number of lawyers on the provision of legal services decreases, because the latter in these cities is relatively high and uniform. Regional centers typically have relatively high availability of legal services. Moscow and Saint Petersburg are characterized by the highest number of law firms and high provision of legal services. Empirical data suggests that there may be a saturation limit beyond which the increase in the number of law firms does not have a substantial impact on provision of legal servicies.

The overwhelming majority (about 85%) of cities is characterized both by low number of law firms and low level of legal services provision. There are two groups of cities by their way of achieving high levels of legal services provision. The first group is composed of cities with large number of law firms. The second group, on the opposite, have small yet number of law firms that nevertheless satisfy the needs of local communitues. Moscow and its region is characterized by relatively high level of legal services availability, as well as Central and Volga regions of Russia in general. Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, and Novosibirsk also have good provision to legal services. On the contrary, in many regions of Far East, North and Caucasus the level of provision of legal services is low.

>>>>

This paper empirically examines the relationship between the number of law firms in Russian cities and the provision of legal services. The existing literature on the availability of services typically focuses on medical services [@Joseph1982; @vanDoorslaer2006; @Weiss2020; @Maleki2024], with relatively little attention paid to legal services. Nevertheless, there is a growing recognition that spatial aspects of legal services, though understudied by geographers, are worthy of investigation [@Blacksell1988; @Patel2008]. This is because they are directly related to the rule of law, the protection of human rights, and the support of social order and stability. The analysis is based on the open data about small and medium-sized businesses (SMB) published by the Federal Tax Service (FTS) of Russia. The primary research question is how the specific provision of legal services is influenced by the number of lawyers (employees at law firms). Additionally, the study aims to identify geographical patterns in the spatial distribution of legal services provision.

@@

**Data.** This research uses a geocoded dataset on small and medium-sized law firms incorporated in Russia. The dataset is derived from FTS open data dumps (archives) of the state SMB Registry[@SMBRegistry2024] and FTS open data on the average number of employees of organizations[@FTSEmployees2024], which were processed using the Python command-line tool *ru-smb-companies* developed by the author of the paper. The dataset contains information about distinct companies, including their names, taxpayer IDs, regions and settlements of incorporation, and main activity codes.

Law firms were selected based on a main activity code equal of 69.10. This selection technique was derived from the analytical paper on the Russian legal services market prepared by the members of the Institute for the Rule of Law [@Moiseeva2016]. Furthermore, the dataset was subjected to the following filters: the year was set to 2021, the company was identified as an organization rather than a sole entrepreneur, the company had declared non-zero revenue and expenditure for the target year (implying that the company was active), and the company's address of registration was in a city rather than in a rural area. In addition to the main dataset, a data on city population was utilized from a freely distributed commercial source [@HFLabs2021].

During the exploratory data analysis, six cities were identified as outliers and subsequently excluded from the dataset. Two of the cities, Moscow and Saint Petersburg, were removed due to the extremely high number of employees at law firms: 18,866 and 5,550, respectively. In contrast, the next city had a considerably smaller number of employees, with less than 2,000. The remaining four cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) are distinguished by a markedly elevated proportion of law firms' employees in their respective populations. This proportion ranges from 38 per 10,000 in Krasnoarmeisk to 417 per 10,000 in Innopolis, in stark contrast to less than 20 per 10,000 in all other cities. After applying all the filters and removing the outliers, the dataset contain information about 576 cities.

**Methods.** The primary analytical technique employed in this study is linear regression. The independent variable in the regression is the number of lawyers (employees at law firms) in a city, which is denoted as *n*. In theory, either the number of law firms or their employees may be used as the independent variable, but law firms in different cities may have different sizes, and thus the number of employees seems to be a more stable and consistent metric. The value for each city is calculated as the sum of the number of employees of each law firm incorporated in that city. If a law firm reports a zero count of employees, it is assumed that the number of employees is equal to one. The dependent variable in the regression is the specific provision of legal services which is denoted as *p* and calculated by dividing the number of employees at law firms in the city by the city's population and multiplying the result by 10,000. In other words, it is the number of lawyers per 10,000 of population. This formula represents a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016].

Cities were classified according to their population size. The population-based categorization is conducted in accordance with the classification system proposed by @Lappo1997. This is done to ascertain whether the hypothesized relationship is dependent on the city size. Linear models were constructed for the entire dataset, without accounting for the groups, or within each group separately. If necessary, the variables were log10-transformed.  The results are then presented in graphical form on the plots. The analysis was conducted in the R programming language in conjunction with additional packages. The source code for this paper can be accessed via the public repository: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability.

**Theoretical model.** Although the modelling described earlier appears straightforward, it is based on a theoretical mathematical framework, and several potential outcomes can be identified before analyzing the data. It is known that a simple linear model is described by the two coefficients: the intercept, which is a term of the linear equation showing the relative shift of the regression line along the y-axis, and the slope, which is a coefficient of the independent variable showing the angle between the regression line and the x-axis. The slope can be either positive, negative, or zero. As a special edge case, it can be also infinite (that is, the regression line is vertical), although, strictly speaking, such a dependency is not a mathematical function. Consequently, the linear fit previously described can yield a result of one of four types, each corresponding to one of the slope options and, subsequently, the pattern of relationship between the number of lawyers and the provision of legal services. These types are illustrated in Figure 1.

```{r figure-1-theoretical-models, fig.cap="**Figure 1** — Four plots showing four types of the theoretical relationship between the number of lawyers and the provision of legal services in cities", fig.height=3}
theoretical_models_plot
```

In the context of regression analysis, Type 1 corresponds to the zero slope. In this instance, the regression line is horizontal. From the standpoint of legal services provision, it describes the absolute equality in the delivery of legal services across cities and its independence of the number of lawyers. This type can be also described as optimal, as it ensures equitable access to legal services (at least in spatial terms). Type 4, similarly to Type 1, also indicates the independence between the number of lawyers and the provision of legal services. However, in this case the number of employees at law firms is fixed, rather than their proportion in the population.

Type 2 is defined by the positive slope and, consequently, positive linear relationship between the number of law firms and the provision of legal services. It appears that in a free market, this type is the most probable, although the specific value of the slope may vary. In such a scenario, the greater the number of lawyers, the easier it is to find a lawyer, and vice versa. In contrast to Type 2, Type 3 exhibits a negative regression slope and a negative relationship between the number of lawyers and the provision of legal services. It is possible that this type may emerge when there is an unexpected and rapid migration of people from the smallest cities to the largest with a nearly uniform (Type 1) distribution of lawyers as a prerequisite. In such circumstances, the specific provision of legal services in the smallest cities increases at a high rate due to the population loss. However, in the largest cities the number of lawyers cannot grow as fast as the population, leading to a decline in the specific provision of legal services. As a result, the slope of the regression line becomes negative.

@@

Figure 2 illustrates the relationship between the number of lawyers and the provision of legal services. The overall connection between the number of law firms and the provision of legal services appears to be nonlinear, with a square root function providing an accurate fit. However, the same relationship within each population size group is well-fitted by a linear model. The coefficients of these models differ, and the slope declines as the city population increases. While the regression lines for the small and medium cities are close to Type 4 (provision is independent of the number of lawyers due to the small variation in the number of lawyers), the modeled relationship in millionaire cities resembles Type 1 (another option for independence, but due to relatively uniform and equal provision regardless of the number of lawyers). The cities from big to extra-large are characterized by a positive correlation between the number of lawyers and the provision of legal services, which can be classified as Type 2.

@@

```{r fig-2-empl-provision, fig.cap="**Figure 2** — The relationship between the number of lawyers and the provision of legal services in cities", fig.height=5}
ep_plot
```

In general, regional centers have a relatively high number of lawyers and a considerable level of legal services provision. This observation is likely attributable to the phenomenon of regional centralization: economic activity and commercial services are concentrated in the administrative centers of regions, which host the governmental bodies and are often the most populous cities within a region. Moscow and Saint Petersburg, although excluded from the data analysis and shown separately on a subplot, are two notable outliers with an extremely high number of lawyers and a high level of legal services provision. Moscow accounts for approximately one-third of all employees at law firms, while Saint Petersburg accounts for approximately 10%. The position of the corresponding data points relative to the main regression lines suggests that there is a saturation point, at which further increase in the number of lawyers has little effect on the provision of legal services.

The general tendencies observed in Figure 2 should be supplemented by an analysis of the cities distribution shown in Figure 3, which extends Figure 2 and displays the binned count of cities according to the number of lawyers and the provision of legal services. The overwhelming majority of cities exhibit a low level of both legal services provision and law firms development. A mere 18 cities are distinguished by a high number of lawyers, exceeding 500. It is noteworthy that a high number of lawyers is favorable to the legal services provision, as all of these cities are at least at the medium level of provision, and none of them has less than 5 lawyers per 10,000 people. However, cities with a relatively low number of lawyers can reach top performance in legal services provision that is on par with the cities with the highest number of lawyers. Moreover, it appears that such small yet well-served cities are even more prevalent than cities with a high level of both lawyers and provision. The relative location of bin counts in the Figure 2 suggests that, in addition to the dominant bottom left corner, there are two main clusters of cities representing two distinct approaches to achieving a high level of legal services provision. One approach is to increase both the absolute number of lawyers and the relative number of lawyers per population, which is characteristic of the largest cities. A second approach is to have a relatively small yet sufficient number of lawyers, which may be typical of small and medium-sized cities.

```{r figure-3-empl-provision-binned, fig.cap="**Figure 3** — Count and proportion of cities with variuos combinations of number of lawyers and the level of legal services provision"}
ep_binned_plot
```

Figure 4 presents a map of cities with relatively high (more than 5 lawyers per 10,000 people) provision of legal services. The most prominent cluster is formed by Moscow and its satellite cities. Other notable areas with high provision include Krasnodar and Stavropol in the south of Russia, Yekaterinburg and Chelyabinsk in the Urals, and Tuymen and Novosibirsk in Western Siberia. There are several regions with no cities that provide a high level of legal services provision (in particular, the Far East, North, and Caucasus). Conversely, the majority of cities with a relatively high level of legal services provision are located in the Central and Volga economic regions of the country.

@@

```{r fig-4-map, fig.cap="**Figure 4** — Map of legal services provision in Russian cities. Only cities with more than 5 lawyers per 10,000 people are shown", fig.height=4}
map
```

@@

The provision of legal services in cities has a positive relationship with the number of lawyers. The overall trend can be approximated by a square root curve, while the relationship in distinct population size groups is close to linear with varying slope of the regression line. In small cities the increase in the number of lawyers has a minimal impact on the level of legal services provision, given the considerable variation of the dependent variable. In larger cities, the relationship is linear and positive. In the cities with a population above one million, the impact of the growth of the number of lawyers on the provision of legal services is less pronounced, as the latter is already high and uniform in these cities. Regional centers typically have a relatively high availability of legal services. Moscow and Saint Petersburg are distinguished by the highest number of lawyers and the high provision of legal services. Empirical data indicates that there may be a saturation limit beyond which the increase in the number of employees at law firms does not have a substantial impact on the provision of legal services

The majority of cities (approximately 85%) are characterized by a paucity of lawyers and a correspondingly limited level of legal services provision. Two distinct groups of cities can be identified based on their approach to achieving high level of legal services provision. The first group is comprised of cities with large number of lawyers. The second group, in contrast, has small number of lawyers that nevertheless meet the needs of local communities. Moscow and its surrounding region exhibit a relatively high level of legal services provision, as do the Central and Volga economic regions of Russia. Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, and Novosibirsk also have a satisfactory provision of legal services. Conversely, in numerous regions of the Far East, North and Caucasus the provision of legal services is notably limited.
